---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# babeldown

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/ropensci-review-tools/babeldown/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ropensci-review-tools/babeldown/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of babeldown is to support workflows that include automatic translation of Markdown-based R content, through DeepL API.

## Installation and setup

You can install the development version of babeldown from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ropensci-review-tools/babeldown")
```

### API URL

The DeepL API URL depends on your API plan.
babeldown uses the DeepL _free_ API URL by default.
If you use a Pro plan, set the API URL via

```{r}
Sys.setenv("DEEPL_API_URL" = "https://api.deepl.com")
```

### API key

Set your API key via the environment variable `DEEPL_API_KEY`.
You could store it with the keyring package and retrieve it like so:

```r
Sys.setenv(DEEPL_API_KEY = keyring::key_get("deepl"))
```

## Troubleshooting

Getting an HTTP error 456 means you've used up all your API credits.
Use `deepl_usage()` (or the online interface) to get your usage data.

## Examples

### File translation

```{r}
english_lines <- c(
  "## A cool section", "",
  "This is the first paragraph. `system.file()` is cool, right?", "",
  "Another paragraph. I really enjoy developing R packages.", "",
  "Do you enjoy debugging?"
)
file <- withr::local_tempfile()
writeLines(english_lines, file)

out_path <- withr::local_tempfile()

babeldown::deepl_translate(
  path = file,
  out_path = out_path,
  source_lang = "EN",
  target_lang = "ES",
  formality = "less"
)

readLines(out_path)
```

### Glossary creation

```{r}
filename <- system.file("example-es-en.csv", package = "babeldown")

# file contents for info
readr::read_tsv(filename, show_col_types = FALSE)

# create (or update) glossary
babeldown::deepl_upsert_glossary(
  filename,
  glossary_name = "rstats-glosario",
  target_lang = "Spanish",
  source_lang = "English"
)
```

### File translation with glossary

```{r}
file <- withr::local_tempfile()
writeLines(english_lines, file)

out_path <- withr::local_tempfile()

babeldown::deepl_translate(
  path = file,
  out_path = out_path,
  source_lang = "EN",
  target_lang = "ES",
  formality = "less",
  glossary = "rstats-glosario"
)

readLines(out_path)
```
