---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# babeldown

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/ropensci-review-tools/babeldown/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ropensci-review-tools/babeldown/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of babeldown is to support workflows that include automatic translation of Markdown-based R content, through [DeepL API](https://www.deepl.com/en/docs-api/).
With babeldown you can translate: Markdown strings (`babeldown::deepl_translate_markdown_string()`) and Markdown files (`babeldown::deepl_translate()`) in general; Quarto book chapters (`babeldown::deepl_translate_quarto()`) and Hugo blog posts (`babeldown::deepl_translate_hugo()`) in particular.

## Installation and setup

You can install the development version of babeldown from [rOpenSci R-universe](https://ropensci.r-universe.dev/):

``` r
install.packages('babeldown', repos = c('https://ropensci.r-universe.dev', 'https://cloud.r-project.org'))
```

Or from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("ropensci-review-tools/babeldown")
```

### API URL

The DeepL API URL depends on your API plan.
babeldown uses the DeepL _free_ API URL by default.
If you use a Pro plan, set the API URL via

```{r}
Sys.setenv("DEEPL_API_URL" = "https://api.deepl.com")
```

### API key

Set your API key via the environment variable `DEEPL_API_KEY`.
You could store it with the [keyring](https://r-lib.github.io/keyring/index.html) package and retrieve it like so:

```{r}
Sys.setenv(DEEPL_API_KEY = keyring::key_get("deepl"))
```

### Line wrapping

We recommend to start a new line in your Markdown document after each sentence or sentence part (after a comma for instance), for more informative Git diffs.

We also recommend not starting a new line in the middle of something that's between square brackets (in particular references and in-line footnotes) as it would break the way babeldown tries to protect those.

#### RStudio Visual Editor

If you use RStudio Visual Editor, you can choose ["sentence" as line-wrapping option](https://rstudio.github.io/visual-markdown-editing/markdown.html#line-wrapping).

## Troubleshooting

Getting an HTTP error 456 means you've used up all your API credits.
Use `deepl_usage()` (or the online DeepL API interface) to get your usage data.

The DeepL API might mix up some punctuation, for instance:

```{r}
babeldown::deepl_translate_markdown_string(
  "[So _incredibly_ **wonderful**](https://ropensci.org)!",
  source_lang = "EN",
  target_lang = "ES"
)
babeldown::deepl_translate_markdown_string(
  "[So _incredibly_ **wonderful**!](https://ropensci.org)",
  source_lang = "EN",
  target_lang = "ES"
)
babeldown::deepl_translate_markdown_string(
  "[So _incredibly_ **wonderful!**](https://ropensci.org)",
  source_lang = "EN",
  target_lang = "ES"
)
```

## Examples

### Markdown string translation

```{r}
babeldown::deepl_translate_markdown_string(
  "[So _incredibly_ **wonderful**](https://ropensci.org)",
  source_lang = "EN",
  target_lang = "ES"
)
```

### File translation

```{r}
english_lines <- c(
  "## A cool section", "",
  "This is the first paragraph. `system.file()` is cool, right?", "",
  "Another paragraph. I really enjoy developing R packages.", "",
  "Do you enjoy debugging?"
)
file <- withr::local_tempfile()
brio::write_lines(english_lines, file)

out_path <- withr::local_tempfile()

babeldown::deepl_translate(
  path = file,
  out_path = out_path,
  source_lang = "EN",
  target_lang = "ES",
  formality = "less"
)

readLines(out_path)
```

You can also indicate YAML fields to translate, using the `yaml_fields` argument.
By default, `title` and `description` are translated.


```{r}
english_lines <- c(
  "---",
  "title: Nice document",
  "description: An example for sure",
  "---",
  "## A cool section", "",
  "This is the first paragraph. `system.file()` is cool, right?", "",
  "Another paragraph. I really enjoy developing R packages.", "",
  "Do you enjoy debugging?"
)
file <- withr::local_tempfile()
brio::write_lines(english_lines, file)

out_path <- withr::local_tempfile()

babeldown::deepl_translate(
  path = file,
  out_path = out_path,
  source_lang = "EN",
  target_lang = "ES",
  formality = "less"
)

readLines(out_path)
```


### Glossary creation or update

```{r}
filename <- system.file("example-es-en.csv", package = "babeldown")

# file contents for info
readr::read_csv(filename, show_col_types = FALSE)

# create (or update) glossary
babeldown::deepl_upsert_glossary(
  filename,
  glossary_name = "rstats-glosario",
  target_lang = "Spanish",
  source_lang = "English"
)
```

### File translation with glossary

```{r}
file <- withr::local_tempfile()
brio::write_lines(english_lines, file)

out_path <- withr::local_tempfile()

babeldown::deepl_translate(
  path = file,
  out_path = out_path,
  source_lang = "EN",
  target_lang = "ES",
  formality = "less",
  glossary = "rstats-glosario"
)

readLines(out_path)
```

### Quarto book chapters

You can use babeldown to translate chapters of a Quarto multilingual book set up with babelquarto.
See the article describing [the workflow](https://docs.ropensci.org/babeldown/articles/quarto.html)

### Hugo posts

You can also use babeldown to translate blog posts of a Hugo multilingual website using leaf bundles (posts as `index.md` inside a content subfolder) and saving translations along each other (for instance Spanish post as `index.es.md` inside the same subfolder).
See `babeldown::deepl_translate_hugo()`.

If your Hugo website is set up differently, either open an issue or use `babeldown::deepl_translate()`.

#### Hugo shortcodes

Hugo shortcodes are supported but not very flexibly: you need to use `param="value"` with no space, and double quotes.
